---
title: "DoReMiTra"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{DoReMiTra User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(SummarizedExperiment)
library(ExperimentHub)
library(DoReMiTra)
```


# Introduction

`DoReMiTra` provides a curated resource of radiation transcriptomic datasets with harmonized metadata for cross-study comparison such as:
  
Radiation_type
Dose
Time_point
Sex
Organism
Exp_setting

The package delivers data as `SummarizedExperiment` objects directly accessible via `ExperimentHub`, enabling easy subsetting, filtering, and comparative analyses.

This vignette guides you through:

* Exploring available datasets
* Loading and summarizing datasets
* Filtering datasets
* Comparing metadata across datasets

# List All Available Datasets

```{r list-datasets, eval=FALSE}
datasets <- list_datasets()
head(datasets)
```

# Search for Specific Datasets

You can filter datasets based on radiation type, organism, and experimental setting.

```{r search-datasets, eval=FALSE}
search_datasets(radiation_type = "x-ray", organism = "Homo sapiens", exp_setting = "ExVivo")
```

# Load a Dataset

To access an individual dataset, use its exact name as provided by `list_datasets()`:

```{r load-dataset, eval=FALSE}
dataset_name <- "SE_Amundson_2008_ExVivo_GSE8917_GPL1708"
se <- get_radiation_data(dataset_name)
se
```

# 5. Summarize a Dataset

You can get a quick metadata summary of a dataset using `se_summary()`:

```{r summarize, eval=FALSE}
se_summary("SE_Amundson_2008_ExVivo_GSE8917_GPL1708")
```

# 6. Compare Metadata Across Multiple Datasets

This is useful for checking compatibility before combined analysis:

```{r compare, eval=FALSE}
se1 <- get_radiation_data("SE_Amundson_2008_ExVivo_GSE8917_GPL1708")
se2 <- get_radiation_data("SE_Park_2017_ExVivo_GSE102971_GPL10332")
compare_datasets(Amundson = se1, Park = se2)
```

# Troubleshooting and Tips

* Always use exact dataset names from `list_datasets()` when calling `get_radiation_data()`.
* Use `search_datasets()` to dynamically find datasets of interest based on key metadata fields.
* If you get a missing dataset error, check for typos or mismatched casing.

# Citation

To cite the datasets or studies, please refer to the original GEO accession IDs (e.g., GSE124612). More details are included in the metadata.

# Session Info

```{r session}
sessionInfo()
```
