---
title: "DoReMiTra"
output: 
  BiocStyle::html_document:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{DoReMiTra User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(SummarizedExperiment)
library(ExperimentHub)
library(DoReMiTra)
```


# Introduction

`DoReMiTra` provides a curated resource of radiation transcriptomic datasets with harmonized metadata for cross-study comparison such as:
  
Radiation_type
Dose
Time_point
Sex
Organism
Exp_setting

The package delivers data as `SummarizedExperiment` objects directly accessible via `ExperimentHub`, enabling easy subsetting, filtering, and comparative analyses.

This vignette guides you through:

* Exploring available datasets
* Loading and summarizing datasets
* Filtering datasets
* Comparing metadata across datasets

# List All Available Datasets

```{r list-datasets}
datasets <- list_DoReMiTra_datasets()
DT::datatable(datasets)
```

# Load All Datasets

You can use the function load_doremitra_datasets() to download all available DoReMiTra datasets from ExperimentHub and assign each as a SummarizedExperiment object in your global environment with its dataset name

```{r load-all-datasets, eval= FALSE}

all_SEs<- get_all_DoReMiTra_datasets()

# Now you can access any loaded dataset by its name, for example:
all_SEs[[SE_Salah_2025_ExVivo]]
```

# Search for Specific Datasets

You can filter datasets based on radiation type, organism, and experimental setting.

```{r search-datasets}
search_DoReMiTra_datasets(radiation_type = "x-ray", 
                          organism = "Homo sapiens", 
                          exp_setting = "ExVivo")
```

# Load a Dataset

To access an individual dataset, use its exact name as provided by `list_datasets()`
This function contains "gene_symbol" argument that is set to TRUE by default, which assigns the gene symbol from the rowData to the rownames of the SE object. if a gene symbol was found to be duplicated, the gene symbol and its corresponding probe id will be appended and assigned to the rownames. if a gene symbol is messing- NA- the probe id will be used instead. otherwise, if the gene_symbol is set to FALSE, probe ids will be used as rownames

```{r load-dataset}
dataset_name <- "SE_Amundson_2008_ExVivo_GSE8917_GPL1708"
se <- get_DoReMiTra_data(dataset_name, gene_symbol = TRUE)
se
```

# Exploring a Dataset

Each dataset is a SummarizedExperiment object with gene-level expression values, sample-level metadata (colData) and gene annotations (rowData):

```{r explore-dataset}
assay(se)[1:5, 1:5] # expression matrix
head(rowData(se)|> as.data.frame()) # gene info
head(colData(se) |> as.data.frame()) # sample info
```

# Summarize a Dataset

You can get a quick metadata summary of a dataset using `se_summary()`:

```{r summarize}
summarize_DoReMiTra_se(se)
```

# Compare Metadata Across Multiple Datasets

This is useful for checking compatibility before combined analysis:

```{r compare}
se1 <- get_DoReMiTra_data("SE_Amundson_2008_ExVivo_GSE8917_GPL1708")
se2 <- get_DoReMiTra_data("SE_Paul_2013_ExVivo_GSE44201_GPL6480")
se_list<- list(Amundson = se1, Paul= se2)
compare_DoReMiTra_datasets(se_list = se_list) |> DT::datatable()
```

# Troubleshooting and Tips

* Always use exact dataset names from `list_DoReMiTra_datasets()` when calling `get_DoReMiTra_data()`.
* Use `search_DoReMiTra_datasets()` to dynamically find datasets of interest based on key metadata fields.
* If you get a missing dataset error, check for typos or mismatched casing.

# Citation

To cite the datasets or studies, please refer to the original GEO accession IDs (e.g., GSE124612). More details are included in the metadata.

# Session Info

```{r session}
sessionInfo()
```
