---
title: "DoReMiTra"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{DoReMiTra User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The DoReMiTra package provides curated transcriptomic datasets derived primarily from public GEO studies that investigate the effects of radiation exposure in both human and mouse samples.

This resource includes:

1- Different radiation types: X-ray, Gamma, Neutron

2- Multiple organisms: Homo sapiens, Mus musculus, Macaca mulatta

3- Experimental contexts: in vitro, in vivo

4- Rich, standardized **SummarizedExperiment** objects

5- Harmonized metadata for cross-study comparisons

These datasets are hosted through ExperimentHub.

# Accessing the Data

You can access the datasets using the **ExperimentHub** interface:

```{r}
library(ExperimentHub)
eh <- ExperimentHub()


query(eh, "DoReMiTra")
```


Once available on ExperimentHub, you can retrieve individual datasets using their EH IDs:

```{r}
se <- eh[["EH1234"]]
```

# Exploring a Dataset

Each dataset is a SummarizedExperiment object with gene-level expression values, sample-level metadata (colData) and gene annotations (rowData):
```{r, eval=FALSE}
assay(se)[1:5, 1:5] # expression matrix
head(rowData(se)) # gene info
head(colData(se)) # sample info
```

# Metadata Overview

The colData of each SE object contains curated metadata, including:

Dose: radiation dose (e.g. "0Gy", "4Gy")

Time_point: time after exposure (e.g. "6hr", "24hr")

Sex, Genotype, Tissue, Cell_type: where applicable

Radiation_type: e.g. "X-ray", "gamma", "Neutron"

Exp_setting: either "vitro" or "vivo"

Organism: scientific name of the source organism

This metadata enables users to subset or filter datasets for comparative analysis.

# Citation

To cite the datasets or studies, please refer to the original GEO accession IDs (e.g., GSE124612). More details are included in the metadata.

# Session Info

```{r}
sessionInfo()
```
