---
title: "DoReMiTra"
output: 
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{DoReMiTra}
  %\VignettePackage{DoReMiTra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: DoReMiTra.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(SummarizedExperiment)
library(ExperimentHub)
library(DoReMiTra)
```


# Introduction

`DoReMiTra` provides a curated resource of radiation transcriptomic datasets with harmonized metadata for cross-study comparison such as:

* `Radiation_type`, either Xray, gamma ray, neutron
* `Dose`, which is commonly measured in Gy
* `Time_point`, normally indicating the time post radiation exposure
* `Sex` of the donors/samples collected
* `Organism`, which at the current state of things is Homo sapiens and Mus musculus
* `Exp_setting` exvivo invivo

The package delivers data as `SummarizedExperiment` objects directly accessible via `ExperimentHub`, enabling easy subsetting, filtering, and comparative analyses.

This vignette guides you through:

* Exploring available datasets
* Loading and summarizing datasets
* Filtering datasets
* Comparing metadata across datasets

# Getting started {#gettingstarted}

To install this package, start R and enter:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("DoReMiTra")
```

Once installed, the package can be loaded and attached to your current workspace
as follows:

```{r loadlib}
library("DoReMiTra")
```

# List All Available Datasets

```{r list-datasets}
datasets <- list_DoReMiTra_datasets()
knitr::kable(datasets)
```

# Load All Datasets

You can use the function `get_all_DoReMiTra_datasets()` to download all available DoReMiTra datasets from ExperimentHub.
In this example, we fetched all the DoReMiTra datasets which is outputted as a named list from which you can easily access any SE object like [@Salah2025] dataset simply by subsetting.

```{r load-all-datasets, eval= FALSE}

all_SEs <- get_all_DoReMiTra_datasets()

# Now you can access any loaded dataset by its name, for example:
all_SEs[["SE_Salah_2025_ExVivo"]]
```

# Search for Specific Datasets

With the `search_DoReMiTra_datasets()` function, you can filter datasets based on radiation type, organism, and experimental setting.

For example, in the following chunk we search for datasets where human samples have been exposed to X-rays in an *ex vivo* setting.

```{r search-datasets}
search_DoReMiTra_datasets(radiation_type = "gamma ray", 
                          organism = "Homo sapiens", 
                          exp_setting = "ExVivo")
```

# Load a Dataset

To access an individual dataset, use its exact name as provided by `list_DoReMiTra_datasets()`.
This function contains "gene_symbol" argument that is set to TRUE by default, which assigns the gene symbol from the `rowData` to the `rownames` of the SE object. if a gene symbol was found to be duplicated, the gene symbol and its corresponding probe id will be appended and assigned to the rownames. if a gene symbol is messing- NA- the probe id will be used instead. otherwise, if the gene_symbol is set to FALSE, probe ids will be used as rownames

```{r load-dataset}
search_DoReMiTra_datasets(radiation_type = "gamma ray", 
                          organism = "Homo sapiens", 
                          exp_setting = "ExVivo")

dataset_name <- "SE_Amundson_2008_ExVivo_GSE8917_GPL1708"
se <- get_DoReMiTra_data(dataset_name, gene_symbol = TRUE)
se
```

# Exploring a Dataset

Each dataset is a SummarizedExperiment object with gene-level expression values, sample-level metadata (`colData`) and gene annotations (`rowData`):

```{r explore-dataset}
assay(se)[1:5, 1:5] # expression matrix
head(rowData(se)) # gene info
head(colData(se)) # sample info
```

# Summarize a Dataset

You can get a quick metadata summary of a dataset using `summarize_DoReMiTra_se()`which will output key information like the author of the study, number of the samples, organism, experiment setting, and radiation type and dose, and a link directing you to the the study

```{r summarize}
summarize_DoReMiTra_se(se)
```

# Compare Metadata Across Multiple Datasets

This is useful for checking compatibility before combined analysis - for example, we first searched for studies that involved gamma ray irradiation of homo sapiens samples. from the number of studies that are shown, we want to compare between [@Park2017] and [@Paul2008]

```{r compare}
search_DoReMiTra_datasets(radiation_type = "gamma ray", 
                          organism = "Homo sapiens", 
                          exp_setting = "ExVivo")

se1 <- get_DoReMiTra_data("SE_Park_2017_ExVivo_GSE102971_GPL10332_HomoSapiens")
se2 <- get_DoReMiTra_data("SE_Paul_2013_ExVivo_GSE44201_GPL6480")
se_list<- list(
  Park = se1, 
  Paul = se2
)
compare_DoReMiTra_datasets(se_list = se_list)
```

# Troubleshooting and Tips

* Always use exact dataset names from `list_DoReMiTra_datasets()` when calling `get_DoReMiTra_data()`.
* Use `search_DoReMiTra_datasets()` to dynamically find datasets of interest based on key metadata fields.
* If you get a missing dataset error, check for typos or mismatched casing.

# How to cite data included in DoReMiTra

To cite the datasets or studies, please refer to the original GEO accession IDs (e.g., GSE124612). More details are included in the metadata, which you can easily access via `summarize_DoReMiTra_se()`.

# Session Info {-}

```{r session}
sessionInfo()
```

# References {-}
